Precision calculations of cosmic shear projections

### 2017-02-09 limber2.tex
MK: partially answered CH's questions; updated figures (less noise in FullSph)

\title[Precision calculations for cosmic shear]{Precision calculations for cosmic shear}  %??
%{Precision calculations of the cosmic shear power spectrum projection}

@@

For a spin-2 field, \cite{2000PhRvD..62d3007H} approximates the edth operator by Cartesian derivatives, and approximates
equation~(\ref{eq:sYlm_def}) as
%
\begin{equation}
  \ell^2 \, _{\pm2}\!{\rm Y}_{\ell m}(\theta, \varphi) =  {\rm e}^{{\mp}2 {\rm i} \phi_\ell}
    (\partial_1 \pm \partial_2)^2 \, {\rm Y}_{\ell m}(\theta, \varphi).
  \label{eq:sYlm_def_flat}
\end{equation}
%
The spin pre-factor $\sqrt{(\ell-1) \ell (\ell+1) (\ell+2)}$ is replaced by $\ell^2$ in flat co-ordinates. \ch{@MK - say why this is so.  Is it just that flat by definition means large $\ell$?}
We find for the flat-sky shear power spectrum
%
\begin{align}
  P^\gamma_{ij}(\ell) 
                 = & \frac 2 \pi \, \ell^4 \, \pref^2
                 \int_{0}^{\chi_{\rm lim}} \frac{\rm d \chi}{\chi} \frac{q_i(\chi)}{a(\chi)}
                \int_{0}^{\chi_{\rm lim}} \frac{\rm d \chi^\prime}{\chi^\prime}
                \frac{q_j(\chi^\prime)}{a(\chi^\prime)}
                %\nonumber \\
                %& \times
                \int_0^\infty \frac{{\rm d} k}{k^2} \, P_{\rm m}(k, \chi, \chi^\prime) \,
                {\rm j}_\ell(k \chi) \, {\rm j}_\ell(k \chi^\prime) .
  \label{eq:P_ell_gamma_full_der}
\end{align}
%
with flat-sky pre-factor $p(\ell) = \ell^4$. 
See App.~\ref{sec:schmidt08} and \ref{sec:giannantonio12} for discussions of alternative expressions for the flat-sky power spectrum.

@@

We follow \cite{2008PhRvD..78l3506L} who derive the second-order Limber
expansion for general projections from 3D to 2D scalar fields in the spherical,
all-sky case. First, we use the identity of Bessel functions
%
\begin{equation}
  {\rm j}_\ell(x) = \sqrt{\frac{\pi}{2x}} {\rm J}_{\ell + 1/2}(x)
\end{equation}
%
in equation~(\ref{eq:C_ell_full}), where ${\rm J}_\ell$ is the first-kind Bessel
function of order $\ell$. Next, \cite{2008PhRvD..78l3506L} solve integrals of
the form
%
\begin{equation}
  \int_0^\infty {\rm d} \chi f(\chi) {\rm J}_{\ell + 1/2}(k \chi)
  = \int_0^\infty {\rm d} x k^{-1} f(x/k) {\rm J}_{\ell + 1/2}(x)
  \label{eq:int_dchi}
\end{equation}
%
by performing a Taylor expansion of the function $f$ around $x = k \chi = \nu$, where
the Bessel function has its approximate maximum.

@@

With this approximation, the $k$-independent term factors out from equation~(\ref{eq:f_LA08}), and we define the separated kernel function $f_{\rm s}$ as
%
\begin{equation}
  f_{\rm s}(\chi) = D_+(\chi) a^{-1}(\chi) \chi^{-3/2} q(\chi) \, ,
  \label{eq:f_LA08_s}
\end{equation}
%
\ch{?? and - do you mean which?} does not need to be considered when taking the derivatives with respect to comoving distances.
The second-order Limber term of the shear power spectrum can then be simplified to
%
\begin{align}
  C^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(\frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac {a(\chi)}{2 D_+(\chi)} \chi^{3/2}
    \left[ q^{}_i f^{\prime\prime}_{{\rm s} j} + f^{\prime\prime}_{{\rm s}i} q^{}_j
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:C_ell_limber2_dr_s}
\end{align}
%
We compute the numerical higher derivatives as follows. The functions $f_{\rm
s}(\chi)$ are fit as power laws with index $\approx 1.5$. The fit is
performed between $\chi_{\rm min} = 0.001 {\rm Mpc}/h$ and $\chi_{\rm max} =
500 \mbox{Mpc}/h$. At larger comoving distances the kernel decreases faster
than a power law, so we exclude this range from the fit. Even though the
derivaties are larger due to the steeper decline, the associated errors are
very small as these scales are down-weighed by the kernel function $f_{\rm
s}$ itself. At $\chi = 500 \mbox{Mpc}/h$ the filter function is less than
$10^{-4}$ of its value of unity at Mpc$/h$.

@@

The extended flat-sky Limber approximation is readily derived from the
spherical case, by replacing the pre-factor $\ellbar^2(\ell, 2)$ with $\ell^4$,
%
\begin{align}
  P^\gamma_{{\rm L1}, ij}(\ell) = & \, \frac{\ell^4}{\nu^4} \, \pref^2 \int {\rm d} \chi \frac{ q_i(\chi) q_j(\chi) }{a^2(\chi)}
  P_{\rm m}\left(k = \frac{\nu}{\chi}; \chi\right);
  \label{eq:P_ell_limber1}
  \\
    P^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ell^4}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(k = \frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac{a(\chi)}{2 D^{-1}_+(\chi)} \chi^{3/2} \left[ q^{}_i f^{\prime\prime}_{{\rm s}j} + f^{\prime\prime}_{{\rm s}i} q^{}_j  
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:P_ell_limber2}
\end{align}
%
Further approximations are commonly made for the pre-factor $p(\ell) = \ell^4/\nu^4$ and $\nu(\ell)$:
%
\begin{enumerate}
  \item $p(\ell) = 1$, this corresponds to $\nu(\ell) = \ell$, which is the \emph{standard} Limber approximation.
    Until recently, i.e.~for all pre-2014 CFHTLenS results, this was the approximation of choice.
  \item $p(\ell) = \ell^4/(\ell + 1/2)^4$. This corresponds to the \emph{extended} Limber
    approximation with $\nu(\ell) = \ell + 1/2$; however the following case is typically employed:
  \item $p(\ell) = 1$, with $\nu = \ell + 1/2$. This is
    a \emph{hybrid} between
    standard and extended Limber approximation, and was used in \cite{KiDS-450,joudaki/etal:2016}. As is shown
    later, this is a better approximation to the full projection than case (ii).
\end{enumerate}
\ch{@MK - in the second order hybrid case does $\ell^4 / \nu^6$ become $\ell^{-2}$ or 1?.  Would this section be clearer if we had $p(\ell)$ in the equations instead of 
$\ell^4/\nu^4$?}

@@

Considering now the spherical-sky cases, we find that including the spherical pre-factor decreases the difference between the extended Limber approximated cases (ExtL1Sph and ExtL2Sph) and the full spherical solution by a factor of a few for $\ell < 5$.   We find that using the spherical-sky second-order extended Limber approximation (ExtL2Sph) yields percent accuracy down to $\ell > 3$.  The numerical calculation of the second-order extended Limber approximation is a factor of \ch{X} faster than the calculation of the full spherical solution.   We note that the noise seen in the right panel of Fig.~\ref{fig:Cl_cases} is due to numerical noise arising from numerical integration errors in the calculation of 
the full spherical solution when $\ell > 20$.  Increasing numerical precision would increase the calculation time still further, relative to the  second-order extended Limber calculation.

@@

On a sphere, correlation functions cannot be related to the power spectrum by the
Hankel transform in equation~(\ref{eqn:xiGG}), as the spherical power spectrum is
formally not defined for non-integer $\ell$ \citep[see][for alternative
spherical-sky formulae for the two-point correlation
function]{2005PhRvD..72b3516C}. In particular, the spherical Bessel functions
necessary to compute the full projection in equation~(\ref{eq:C_ell_full}) is defined only
for integer order $\ell$. \mk{Is this really true?}

As we have shown in section~\ref{sec:comp}, however, the spherical
second-order extended Limber approximation provides a percent-level precision
representation of the full spherical projection for $\ell > 3$.  
In the Limber approximations the Bessel functions expanded \ch{to by ?? which one??}
polynomials, allowing for arbitrary wave modes $\ell$. The spherical pre-factor
$\ellbar(\ell, 2)$ (equation~ref{eq:ellbar}) can be generalised to non-integer
arguments and is positive for $\ell \le 2$. We can thus use the spherical power
spectrum in the Limber approximations for the Hankel transformation to compute
the two-point correlation functions.  We therefore do not
present predictions for the two-point shear correlation function using the full
spherical projection, instead choosing the spherical second-order extended
Limber approximation (Ext2Sph) as our reference in this case.

@@

\ch{MNRAS rules;  equation when you refer to your equation.  Equation when it is in another paper.}

>>>>

\title[Precision calculations for cosmic shear projections]{Precision calculations of cosmic shear projections}  %??
%{Precision calculations of the cosmic shear power spectrum projection}

@@

For a spin-2 field, \cite{2000PhRvD..62d3007H} approximates the edth operator
by Cartesian derivatives, and approximates equation~(\ref{eq:sYlm_def}) as
%
\begin{equation}
  \ell^2 \, _{\pm2}\!{\rm Y}_{\ell m}(\theta, \varphi) \approx  {\rm e}^{{\mp}2 {\rm i} \phi_\ell}
    (\partial_1 \pm \partial_2)^2 \, {\rm Y}_{\ell m}(\theta, \varphi).
  \label{eq:sYlm_def_flat}
\end{equation}
%
The spin pre-factor $\sqrt{(\ell-1) \ell (\ell+1) (\ell+2)}$ is replaced by
$\ell^2$ in flat co-ordinates, \mk{an approximation that holds for large $\ell$.}
 \ch{@MK - say why this is so.  Is it just that
flat by definition means large $\ell$?} \mk{I think it's a large-ell approximation,
but I don't know whether this is an assumption to derive this approximation, or whether it follows from that.
I'll have to check this.}
We find for the flat-sky shear power
spectrum
%
\begin{align}
  P^\gamma_{ij}(\ell) 
                 = & \frac 2 \pi \, \ell^4 \, \pref^2
                 \int_{0}^{\chi_{\rm lim}} \frac{\rm d \chi}{\chi} \frac{q_i(\chi)}{a(\chi)}
                \int_{0}^{\chi_{\rm lim}} \frac{\rm d \chi^\prime}{\chi^\prime}
                \frac{q_j(\chi^\prime)}{a(\chi^\prime)}
                %\nonumber \\
                %& \times
                \int_0^\infty \frac{{\rm d} k}{k^2} \, P_{\rm m}(k, \chi, \chi^\prime) \,
                {\rm j}_\ell(k \chi) \, {\rm j}_\ell(k \chi^\prime) .
  \label{eq:P_ell_gamma_full_der}
\end{align}
%
with flat-sky pre-factor $p(\ell) = \ell^4$. 
See App.~\ref{sec:schmidt08} and \ref{sec:giannantonio12} for discussions of alternative expressions for the flat-sky power spectrum.

@@

We follow \cite{2008PhRvD..78l3506L} who derive the second-order Limber
expansion for general projections from 3D to 2D scalar fields in the spherical,
all-sky case. First, we use the identity of Bessel functions
%
\begin{equation}
  {\rm j}_\ell(x) = \sqrt{\frac{\pi}{2x}} {\rm J}_{\ell + 1/2}(x)
  \label{eq:jJ}
\end{equation}
%
in equation~(\ref{eq:C_ell_full}), where ${\rm J}_\ell$ is the first-kind Bessel
function of order $\ell$. Next, \cite{2008PhRvD..78l3506L} solve integrals of
the form
%
\begin{equation}
  \int_0^\infty {\rm d} \chi f(\chi) {\rm J}_{\ell + 1/2}(k \chi)
  = \int_0^\infty {\rm d} x k^{-1} f(x/k) {\rm J}_{\ell + 1/2}(x)
  \label{eq:int_dchi}
\end{equation}
%
by performing a Taylor expansion of the function $f$ around $x = k \chi = \nu$, where
the Bessel function has its approximate maximum.

@@

With this approximation, the \mk{$\chi$-independent term $P_{\rm m}(k)$}
factors out from equation~(\ref{eq:f_LA08}), and we define the separated kernel
function $f_{\rm s}$ as
%
\begin{equation}
  f_{\rm s}(\chi) = D_+(\chi) a^{-1}(\chi) \chi^{-3/2} q(\chi) \,
  \label{eq:f_LA08_s}
\end{equation}
%
The second-order Limber term of the shear power spectrum can then be simplified to
%
\begin{align}
  C^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(\frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac {a(\chi)}{2 D_+(\chi)} \chi^{3/2}
    \left[ q^{}_i f^{\prime\prime}_{{\rm s} j} + f^{\prime\prime}_{{\rm s}i} q^{}_j
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:C_ell_limber2_dr_s}
\end{align}
%
We compute the numerical higher derivatives as follows. The functions $f_{\rm
s}(\chi)$ are fit as power laws with index $\approx 1.5$. The fit is
performed between $\chi_{\rm min} = 0.001 {\rm Mpc}/h$ and $\chi_{\rm max} =
500 \mbox{Mpc}/h$. At larger comoving distances the kernel decreases faster
than a power law, so we exclude this range from the fit. Even though the
derivaties are larger due to the steeper decline, the associated errors are
very small as these scales are down-weighed by the kernel function $f_{\rm
s}$ itself. At $\chi = 500 \mbox{Mpc}/h$ the filter function is less than
$10^{-4}$ of its value of unity at Mpc$/h$.

@@

The extended flat-sky Limber approximation is readily derived from the
spherical case, by replacing the pre-factor $\ellbar^2(\ell, 2)$ with $\ell^4$,
%
\begin{align}
  P^\gamma_{{\rm L1}, ij}(\ell) = & \, \frac{\ell^4}{\nu^4} \, \pref^2 \int {\rm d} \chi \frac{ q_i(\chi) q_j(\chi) }{a^2(\chi)}
  P_{\rm m}\left(k = \frac{\nu}{\chi}; \chi\right);
  \label{eq:P_ell_limber1}
  \\
    P^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ell^4}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(k = \frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac{a(\chi)}{2 D^{-1}_+(\chi)} \chi^{3/2} \left[ q^{}_i f^{\prime\prime}_{{\rm s}j} + f^{\prime\prime}_{{\rm s}i} q^{}_j  
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:P_ell_limber2}
\end{align}
%
Further approximations are commonly made for the pre-factor $p(\ell) = \ell^4/\nu^4$ and $\nu(\ell)$:
%
\begin{enumerate}
  \item $p(\ell) = 1$, this corresponds to $\nu(\ell) = \ell$, which is the \emph{standard} Limber approximation.
    Until recently, i.e.~for all pre-2014 CFHTLenS results, this was the approximation of choice.
  \item $p(\ell) = \ell^4/(\ell + 1/2)^4$. This corresponds to the \emph{extended} Limber
    approximation with $\nu(\ell) = \ell + 1/2$; however the following case is typically employed:
  \item $p(\ell) = 1$, with $\nu = \ell + 1/2$. This is
    a \emph{hybrid} between
    standard and extended Limber approximation, and was used in \cite{KiDS-450,joudaki/etal:2016}. As is shown
    later, this is a better approximation to the full projection than case (ii).
\end{enumerate}
%
\ch{Would this section be clearer if we had $p(\ell)$ in the equations instead of 
$\ell^4/\nu^4$?}

@@

Considering now the spherical-sky cases, we find that including the spherical
pre-factor decreases the difference between the extended Limber approximated
cases (ExtL1Sph and ExtL2Sph) and the full spherical solution by a factor of a
few for $\ell < 5$.   We find that using the spherical-sky second-order
extended Limber approximation (ExtL2Sph) yields percent accuracy down to $\ell
> 3$.  The numerical calculation of the second-order extended Limber
approximation is a factor of \mk{15 times} faster than the calculation of the
full spherical solution \mk{for the first $18$ $\ell$-modes.} We note that the
\mk{sub-$0.1\%$}-noise seen in the right panel of Fig.~\ref{fig:Cl_cases} is
due to numerical noise arising from numerical integration errors in the
calculation of the full spherical solution when $\ell > 20$.

% Compuation times:
% l>=20, 20 modes
% full 37s
% L2   2.14s
% L1   0.8s

@@

On a sphere, correlation functions cannot be related to the power spectrum by
the Hankel transform in equation~(\ref{eqn:xiGG}), as the spherical power
spectrum is formally not defined for non-integer $\ell$ \citep[see][for
alternative spherical-sky formulae for the two-point correlation
function]{2005PhRvD..72b3516C}. In particular, the spherical Bessel functions
necessary to compute the full projection in equation~(\ref{eq:C_ell_full}) is
defined only for integer order $\ell$. \mk{I think this is actually not true,
since J$_\nu$ can be computed for fractional $\nu$, and then one could use the
identity (\ref{P_kappa_full_l1}). Or is it?}.

As we have shown in section~\ref{sec:comp}, however, the spherical
second-order extended Limber approximation provides a percent-level precision
representation of the full spherical projection for $\ell > 3$.  
In the Limber approximations the Bessel functions \mk{are expanded into}
polynomials, allowing for arbitrary wave modes $\ell$. The spherical pre-factor
$\ellbar(\ell, 2)$ (equation~ref{eq:ellbar}) can be generalised to non-integer
arguments and is positive for $\ell \le 2$. We can thus use the spherical power
spectrum in the Limber approximations for the Hankel transformation to compute
the two-point correlation functions.  We therefore do not
present predictions for the two-point shear correlation function using the full
spherical projection, instead choosing the spherical second-order extended
Limber approximation (Ext2Sph) as our reference in this case.

@@
